A Pilot Multicenter Cluster Randomized Trial to Compare The Effect of Trauma Life Support Training Programs on Patient and Provider Outcomes

### 2024-02-02 pilot-publication/manuscript.Rmd

---
## compile with:
## Rscript -e "rmarkdown::render('manuscript.Rmd', output_format = 'all')"
title: A Pilot Multicenter Cluster Randomized Trial to Compare The Effect of Trauma Life Support Training Programs on Patient and Provider Outcomes  
author: Trauma life support training Effectiveness Research Network (TERN) collaborators
bibliography: bibliography.bib
csl: nature.csl
output:
    bookdown::pdf_document2:
      toc: false
      number_sections: false
##    md_document: default
##    bookdown::word_document2:
##      toc: false
##      number_sections: false
##      reference_docx: "reference.docx"
---

@@

**Introduction** Trauma life support training programmes aim to improve trauma outcomes but there is no evidence from controlled trials to show that they work. We conducted a pilot study to assess the feasibility of conducting a cluster randomised controlled trial comparing the effect of Advanced Trauma Life Support^®^ (ATLS^®^) and Primary Trauma Care (PTC) with standard care on patient outcomes. **Methods and analysis** We piloted a pragmatic three-armed parallel, cluster randomised, controlled trial in India, where neither of these programmes are routinely taught. We recruited tertiary hospitals and included trauma patients and residents managing these patients. Two hospitals were randomised to ATLS^®^, two to PTC, and three to standard care. The primary outcome was all cause mortality at 30 days from the time of arrival to the emergency department. We conducted community consultations in parallel with the pilot trial. **Ethics and dissemination** We obtained ethical approval from all participating centres. **Results** Between `r start.date` and `r end.date` we included `r n.patients` patients and `r n.residents` residents from `r n.centres` centres. The percentage of patients who consented to follow up was `r p.consent`% and the resident recruitment rate was `r recruitment.rate.residents`%. The lost to follow up rate was `r rate.lost.to.follow.up`%. `r stringr::str_to_sentence(english::english(m30d.standard.care))` patients died within 30 days in the standard care arm, `r m30d.atls` patients in the ATLS^®^ arm, and `r m30d.ptc` patients in the PTC arm. **Conclusions** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS^®^, PTC, and standard care on patient outcomes should be feasible after incorporating key lessons from this pilot.

@@

### Patient Participants
Adults (15 years or older) who presented to the emergency department at participating hospitals with a history of trauma. History of trauma is here defined as having any of the external causes of morbidity and mortality listed in block V01-Y36, chapter XX of the International Classification of Disease version 10 (ICD-10) codebook as reason for presenting. We explored intervention effects across the following clinical subgroups: men, women, blunt multisystem trauma, penetrating trauma, shock, severe traumatic brain injury, and elderly, as defined by Hornor et al [@Hornor2018]. The consent form for patients are available as Supplemental Material 1.

@@

Consent was sought from the residents in each of the intervention groups before they underwent the ATLS^®^ or PTC training. The consent form for residents is available as Supplemental Material 2. We did not ask for consent from residents at the units in the control hospitals as their practice was not affected by this pilot and we did not collect any personal identifiable data on them.

@@

The primary participant outcome was all cause mortality within 30 days from the time of arrival to the emergency department. The primary outcome and most secondary outcome were assessed and compared both as final values and as change from baseline. All outcomes that pertain to the individual participant level are detailed in Supplemental Material 3. We decided to include a large number of outcomes,
including some more exploratory, so that we could test their feasibility and relevance.

@@

## Variables
The project officers collected data on demographics, time of injury to arrival at the participating hospital, time to recording vital signs, vital signs, and times to and management details including imaging and surgery. Details of any injury sustained were collected and coded using ICD 10 and the Abbreviated Injury Scale (AIS). For ICD 10 coders underwent the WHO online ICD 10 training module and for AIS they were accredited. Based on AIS we calculated the Injury Severity Score (ISS) and the New ISS (NISS). Supplemental Material 4 contains a full variable list, with definitions.

@@

# Results
```{r}
females <- inline_text(overall.sample.characteristics.table,
  variable = patinfo__pt_gender,
  level = "Female"
)
median.age <- inline_text(overall.sample.characteristics.table,
  variable = patinfo__pt_age
)
median.iss <- inline_text(overall.sample.characteristics.table,
  variable = riss
)
median.niss <- inline_text(overall.sample.characteristics.table,
  variable = niss
)
m30d <- inline_text(overall.sample.characteristics.table,
  variable = outcomes__alive_after_30_days,
  level = "Yes"
)
```

@@

The study flowchart is shown in Figure X and patient sample
characteristics across trial arms are shown in Table
\@ref(tab:sample-characteristics). Overall, the number of females were `r females`, the median (IQR) age was `r median.age` years, and the median ISS (IQR) and NISS (IQR) were `r median.iss` and `r median.niss` respectively. A total of `r m30d` patients had the primary outcome of mortality at 30 days after arrival to the emergency department.

@@

## Resident Participant Outcomes
A total of `r n.residents` residents were trained during this
pilot study. Overall their median confidence in managing trauma patients was `r median.confidence` (IQR `r iqr.confidence`) on a 10 point Likert scale with 10 being most confident. <!--In the post training period, the confidence in the ATLS arm was X (IQR X - X), compared to X (IQR X - X) in the PTC arm, and X (IQR X - X) in the standard care arm.-->

## Feasibility Outcomes
<!--Out of X potentially eligible patients, X (X%) patients were enrolled
in the study-->. Out of `r n.eligible.residents` potentially eligible residents, `r n.residents` (`r recruitment.rate.residents`%) participated in the study. The lost to follow up rate was `r rate.lost.to.follow.up`, with `r n.lost.to.follow.up` out of `r n.patients` patients not completing 30 day follow up. Among residents the pass rate, after two attempts, was `r rate.pass`%. The missing data rate ranged from `r min.missing.data` to `r max.missing.data`, with details for selected variables shown in Table \ref@(tab:sample-characteristics). The differences in distributions between observed and extracted data, for selected variables that were collected through observation or interview, are shown in Table \ref@(tab:observed-vs-retrospective). Overall, there were considerably more missing values in extracted data compared to observed data.

@@

```{r secondary-outcomes, echo = FALSE, results = "asis"}
table1::t1kable(secondary.outcomes.table, longtable = TRUE) %>%
  kableExtra::kable_styling(latex_options = c("repeat_header"), font_size = 8)
```

>>>>

---
## compile with:
## Rscript -e "rmarkdown::render('manuscript.Rmd', output_format = 'all')"
title: A Pilot Multicenter Cluster Randomized Trial to Compare The Effect of Trauma Life Support Training Programs on Patient and Provider Outcomes  
author: Trauma life support training Effectiveness Research Network (TERN) collaborators
bibliography: bibliography.bib
csl: nature.csl
output:
    bookdown::pdf_document2:
      toc: false
      number_sections: false
    md_document: default
    bookdown::word_document2:
      toc: false
      number_sections: false
      reference_docx: "reference.docx"
---

@@

**Introduction** Trauma life support training programmes aim to improve trauma outcomes but there is no evidence from controlled trials to show that they work. We conducted a pilot study to assess the feasibility of conducting a cluster randomised controlled trial comparing the effect of Advanced Trauma Life Support^®^ (ATLS^®^) and Primary Trauma Care (PTC) with standard care on patient outcomes. **Methods and analysis** We piloted a pragmatic three-armed parallel, cluster randomised, controlled trial in India, where neither of these programmes are routinely taught. We recruited tertiary hospitals and included trauma patients and residents managing these patients. Two hospitals were randomised to ATLS^®^, two to PTC, and three to standard care. The primary outcome was all cause mortality at 30 days from the time of arrival to the emergency department. We conducted community consultations in parallel with the pilot trial. **Ethics and dissemination** We obtained ethical approval from all participating centres. **Results** Between `r start.date` and `r end.date` we included `r n.patients` patients and `r n.residents` residents from `r n.centres` centres. The percentage of patients who consented to follow up was `r p.consent`% and the resident recruitment rate was `r recruitment.rate.residents`%. The lost to follow up rate was `r rate.lost.to.follow.up`%. `r m30d.standard.care` patients died within 30 days in the standard care arm, `r m30d.atls` patients in the ATLS^®^ arm, and `r m30d.ptc` patients in the PTC arm. **Conclusions** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS^®^, PTC, and standard care on patient outcomes should be feasible after incorporating key lessons from this pilot.

@@

### Patient Participants
Adults (15 years or older) who presented to the emergency department at participating hospitals with a history of trauma. History of trauma is here defined as having any of the external causes of morbidity and mortality listed in block V01-Y36, chapter XX of the International Classification of Disease version 10 (ICD-10) codebook as reason for presenting. We explored intervention effects across the following clinical subgroups: men, women, blunt multisystem trauma, penetrating trauma, shock, severe traumatic brain injury, and elderly, as defined by Hornor et al [@Hornor2018].

@@

Consent was sought from the residents in each of the intervention groups before they underwent the ATLS^®^ or PTC training. We did not ask for consent from residents at the units in the control hospitals as their practice was not affected by this pilot and we did not collect any personal identifiable data on them.

@@

The primary participant outcome was all cause mortality within 30 days from the time of arrival to the emergency department. The primary outcome and most secondary outcome were assessed and compared both as final values and as change from baseline. All outcomes that pertain to the individual participant level are detailed in Supplemental Material 1. We decided to include a large number of outcomes, including some more exploratory, so that we could test their feasibility and relevance.

@@

## Variables
The project officers collected data on demographics, time of injury to arrival at the participating hospital, time to recording vital signs, vital signs, and times to and management details including imaging and surgery. Details of any injury sustained were collected and coded using ICD 10 and the Abbreviated Injury Scale (AIS). For ICD 10 coders underwent the WHO online ICD 10 training module and for AIS they were accredited. Based on AIS we calculated the Injury Severity Score (ISS) and the New ISS (NISS). Supplemental Material 2 contains a full variable list, with definitions.

@@

# Results

@@

The study flow diagram is shown in Figure \@ref(fig:flow-diagram) and patient sample
characteristics across trial arms are shown in Table \@ref(tab:sample-characteristics). Overall, the number of females were `r females`, the median (IQR) age was `r median.age` years, and the median ISS (IQR) was `r median.iss`. A total of `r m30d` patients had the primary outcome of mortality at 30 days after arrival to the emergency department.

```{r flow-diagram, echo = FALSE, fig.cap="Study flow diagram. Abbreviations: ATLS, Advanced Trauma Life Support; PTC, Primary Trauma Care."}
create_flow_diagram()
knitr::include_graphics("flow-diagram.png")
```

```{r sample-characteristics, echo = FALSE}
```

```{r sample-characteristics, echo = FALSE, results = "asis"}
```

@@

## Resident Participant Outcomes
A total of `r n.residents` residents were trained during this
pilot study. Overall their median confidence in managing trauma patients was `r median.confidence` (IQR `r iqr.confidence`) on a 10 point Likert scale with 10 being most confident. In the pre training period, the median confidence in the stadard care arm was `r median.comfort.standard.care.pre` (IQR `r iqr.comfort.standard.care.pre`), compared to `r median.comfort.atls.pre` (IQR `r iqr.comfort.atls.pre`) in the ATLS and `r median.comfort.ptc.pre` (IQR `r iqr.comfort.ptc.pre`) in the PTC arm. In the post training period, the corresponding figures were `r median.comfort.standard.care.post` (IQR `r iqr.comfort.standard.care.post`) in the standard care arm, `r median.comfort.atls.post` (IQR `r iqr.comfort.atls.post`) in the ATLS and `r median.comfort.ptc.post` (IQR `r iqr.comfort.ptc.post`) in the PTC arm.

## Feasibility Outcomes
<!--Out of X potentially eligible patients, X (X%) patients were enrolled
in the study-->. Out of `r n.eligible.residents` potentially eligible residents, `r n.residents` (`r recruitment.rate.residents`%) participated in the study. The lost to follow up rate was `r rate.lost.to.follow.up`, with `r n.lost.to.follow.up` out of `r n.patients` patients not completing 30 day follow up. Among residents the pass rate, after two attempts, was `r pass.rate`%. The missing data rate ranged from `r min.missing.data` to `r max.missing.data`, with details for selected variables shown in Table \ref@(tab:sample-characteristics). The differences in distributions between observed and extracted data, for selected variables that were collected through observation or interview, are shown in Table \ref@(tab:observed-vs-retrospective). Overall, there were considerably more missing values in extracted data compared to observed data.

@@
